---
output:
  md_document:
    variant: gfm
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# demres

The goal of __{demres}__ is to provide easy functions to calculate different time-varying and time-constant demographic resilience metrics. It also allows plotting the resulting metrics and measuring the discrepancy between the time-varying and the time-constant approach. 

The different metrics provided are:   
- Convergence time    
- Damping ratio    
- Reactivity   
- Maximum amplification   
- Maximum attenuation   

It is build around one (direct) dependency:

- [__{popdemo}__](https://github.com/iainmstott/popdemo)

__{demres}__ aims at being compatible with both _tidyverse_ and _base_ R dialects.

## Package installation

You can install this package using __{remotes}__ (or __{devtools}__):

```{r, installation, eval = FALSE}
remotes::install_github("JulieLouvrier/demres")

```

## Basic usage of the package
### Load the package {demres}

```{r, loading pkg}
library(demres)
```

### Import Matrix Population Models 
```{r, import data}
# load data
data(adeliepenguin)
```

### Calculate demographic resilience metrics based on a list of population matrices

The function `resilience` is made to calculate demographic resilience metrics based on a list of matrix projection models. It returns a `dataframe` with both time-varying and time-constant approaches when requested using the argument `time`.

```{r, resilience metrics}
# Create initial vector
set.seed(125435)
vec <- round(runif(2) * 100 )
vector_list <- lapply(1:28, function(i) vec)

#let's calculate all metrics of demographic resilience for with one initial vector
AP_demres <-
   resilience(
     listA = adeliepenguin,
     metrics = "all",
     vector = vector_list,
     TDvector = FALSE,
     popname = "adelie penguin",
     verbose = TRUE,
     return.N = TRUE,
     return.t = TRUE
   )

#let's take a look at what we got
AP_demres
```

### Assess the distance between the time-varying and the time-constant approaches

The function `summary` calculates the variation in the time-varying metrics: 

```{r, variation}
#let's calculate the default variation measures, namely mean and sd 
summary_AP <- summary(AP_demres)

#let's take a look at what we got
summary_AP
```

### Plot the results

Function `plot` provides a plot to visually inspect the resilience metric along a time axis

```{r, fig.width = 6, fig.height = 6}
plot(AP_demres,
     listA = adeliepenguin,
     vector = vector_list,
     timeproj = 5,
     sort = FALSE,
     compare = TRUE)

```


