---
output:
  md_document:
    variant: markdown_github
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# demres

The goal of __{demres}__ is to provide easy functions to calculate different time-varying and time- constant demographic resilience metrics. It also allows plotting the resulting metrics and measuring the discrepancy between the time-varying and the time-constant approach. 

The different metrics provided are:   
- Damping ratio    
- Inertia  
- Reactivity   
- Maximum amplification   
- Maximum attenuation   

It is build around one dependency:

- [__{popdemo}__](https://github.com/r-lib/rlang)

__{demres}__ aims at being compatible with both _tidyverse_ and _base_ R dialects.

## Package installation

You can install this package using __{remotes}__ (or __{devtools}__):

```{r, installation, eval = FALSE}
remotes::install_github("JulieLouvrier/demres")

```

## Basic usage of the package

### Load the package {demres}

```{r, loading pkg}
library(demres)
```

### Import Matrix Population Models 
```{r, import data}
# load data
data(bluecrane)

```

### Calculate demographic resilience metrics based on one Matrix

The function `calc_resilience` is made to calculate demographic resilience metrics based on one Matrix Population Model. It returns a `dataframe`

```{r, resilience metrics}
#select the first matrix 
Crane_mat1 <- bluecrane[[1]]

# Create initial vector
set.seed(1)
Crane_vec1 <- runif(5)
Crane_vec1 <- Crane_vec1 / sum(Crane_vec1) #scales the vec to sum to 1
Crane_vec1 <- c(0, 0, 0, 0, 1)

#let's calculate all metrics of demographic resilience for that one matrix with one initial vector
all_BlueCrane1_demres <- calc_resilience(Crane_mat1, metrics = "all",
vector = Crane_vec1, bounds = TRUE, popname = "Blue Crane")
attr(all_BlueCrane1_demres, "msg")

#let's take a look at what we got
all_BlueCrane1_demres

```


### Calculate demographic resilience metrics based on a list of Matrices

The function `demres` is made to calculate demographic resilience metrics based on a lost Matrix Population Models. It returns a `dataframe` with both time-varying and time-constant approaches when required with the option `time`.

```{r, demres}
#let's calculate all metrics of demographic resilience for that one matrix with one initial vector
all_BlueCrane_demres <- demres(bluecrane, metrics = c("all"),
vector = Crane_vec1, bounds = TRUE, popname = "Blue Crane", time = "both")

#let's take a look at what we got
all_BlueCrane_demres
```

### Assessing the distance between the time-varying and the time-constant approaches

The function `demres_dist` calculates the distance between the time-varying resilience metric and the time-constant one by measuring the RMSE, rRMSE and the MAPE: 

#### RMSE: 
$$\sqrt{mean((TV-TC)^2)}$$   
with TV - the time-varying resilience metric and TC - the time-constant one

#### rRMSE: 
$$\sqrt{mean((TV-TC)^2)} / sd(TV)$$   

#### MAPE: 
$$ \frac{(mean(abs(TV - TC))}{100}$$ 

```{r, distance}
#let's calculate all measures of distance for inertia
dist_BC <- demres_dist(table = all_BlueCrane_demres, metric = "inertia", measure = "all")

#let's take a look at what we got
dist_BC
```

### plotting the results

`demres_plot` provides a plot to visually inspect the resilience metric along a time axis

```{r, plotting}
#specifying the metric that we want to visually inspect
metric <- "inertia"
#plotting
demres_plot(table = all_BlueCrane_demres, metric = metric, plotname = paste0(getwd(), "/plots/plot_demres_", metric, ".pdf"), RMSE = TRUE)

```
