---
title: "demres: quantification of demographic resilience over time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{demres: quantification of demographic resilience over time}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "##  ", collapse = TRUE, fig.align = "center")
library(demres)
options(digits = 4)
```

The goal of __{demres}__ is to provide easy functions to calculate different time-varying demographic resilience metrics. It also allows plotting the resulting metrics and measuring the variability in the time-varying metrics. 

The different metrics provided are:   
- _Convergence time_    
- _Damping ratio_    
- _Maximum amplification_   
- _Maximum attenuation_   
- _Reactivity_   

It is build around one dependency:

- [__{popdemo}__](https://github.com/iainmstott/popdemo)

__{demres}__ aims at being compatible with both _tidyverse_ and _base_ R dialects.

***
# Installing `demres`
__{demres}__ is available on [GitHub](https://github.com/JulieLouvrier/demres). 
Running vignette usually requires the latest version of the package. 

## Package installation from Github

You can install this package using __{remotes}__ (or __{devtools}__):

```{r, installation, eval = FALSE}
# Install dependencies from CRAN:
install.packages(c("devtools", "remotes"))

#Installing from GitHub
remotes::install_github("JulieLouvrier/demres")
```
***  
# Matrix Population Models
## adelie penguin models

We will use a list of matrix projection models (MPMs) of an Adelie penguin population (*Pygoscelis adeliae*) from the Copacabana colony on King George Island, Antarctica, from the study by Hinke et al. (2017): "Variable vital rates and the risk of population declines in Adelie
penguins from the Antarctic Peninsula region" [^1].

The MPMs were extracted from the COMADRE database (https://compadre-db.org/). The data is a list of 28 matrices of dimension 2 x 2. The 2 stage classes have been defined based on age and are the following:     
- "Juveniles" - ages 0–2      
- "Adults" - ages 3+         
   

[^1]: [Hinke et al. 2017 Ecosphere, 8(1), e01666.](https://esajournals.onlinelibrary.wiley.com/doi/epdf/10.1002/ecs2.1666)

```{r echo = FALSE}
penguinpic <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/1/12/Pygoscelis_adeliae.png")
par(mar = c(0, 0, 0, 0))
plot(as.raster(penguinpic))
text(10, 10, adj = c (0, 0), col = "white",
     "Adelie penguin")
```

Load in the data:  
```{r}
data(adeliepenguin); adeliepenguin

```

The elements constituting the matrices are known as vital rates. These represent the survival rates in a given (st)age class, transition probabilities from one (st)age class to another, and fecundity (i.e. per-capita number of offspring that is contributed to the population by each (st)age class), per unit of time. For Adelie penguins, the highest value of fecundity (0.75) was found in year 14 of the study (see `adeliepenguin[[14]]`), while this parameter reached its lowest value of 0.16 in year 7 of the study period (see `adeliepenguin[[7]]`). The adult survival rate varied across years with its highest value (0.98) on years 4, 9, 12 and 15 and lowest values (0.72) on the year 22. The probability to transition from a juvenile to an adult reached its highest value (0.89) on year 21 and its lowest value (0.53) on years 10 and 28. 

##HERE
```{r, data plot}
# Plotting the different demographic parameters

# Fecundity 
adeliepenguin[[1]][1,2]

```

# Assess time-varying demographic resilience 
The function `resilience` is made to calculate demographic resilience metrics based on a list of MPMs. It returns a `dataframe` with time-varying demographic resilience metrics with the option `time = "varying"`.

The different resilience metrics that are calculated[^2] can be :         

- _**Convergence time**_ : Time to convergence of a population to asymptotic growth rate obtained from the model projection.         

- _**Damping ratio**_ : Dimensionless measure of convergence to stable growth. Smaller numbers represent slower convergence. _Damping ratio_ is calculated using 

$$\textit{damping ratio} = {\rho} = \frac{\lambda_1}{||\lambda_2||}$$
With $\lambda_1$ the dominant eigenvalue and $\lambda_2$ the largest subdominant eigenvalue.

- _**Reactivity**_ : The population density that can be reached in the first time step after disturbance. _Reactivity_ is calculated using  

$$\textit{reactivity} = {P_1} = ||\hat{\textbf{A}} \hat{\textbf{n}}_0||_1$$
With **A** the matrix population model and $\hat{\textbf{A}}$ the standardized population matrix which is calculated as $\hat{\textbf{A}} = \frac{\textbf{A}}{\lambda_{1}}$. The vector $\hat{\textbf{n}}_0$ represents the initial demographic distribution, standardized to sum to 1 and is calculated as $\hat{\textbf{n}}_0 = \frac{{\textbf{n}}_0}{||\textbf{n}_0||_1}$. 

- _**Maximum amplification**_ : The largest population density that can be reached at any time after disturbance. This metric is calculated using:
$$maximum\ amplification = \bar{P}_{max} = \max_{t > 0}||\hat{\textbf{A}}^t\hat{\textbf{n}}_0||_1$$

- _**Maximum attenuation**_ : The lowest population density that can be reached at any time after disturbance. This metric is calculated using: 

$$maximum\ attenuation = \underline{P}_{min} = \min_{t > 0}||\hat{\textbf{A}}^t\hat{\textbf{n}}_0||_1$$
_Maximum amplification_ and _maximum attenuation_ can occur at any point along the projection. Some populations only have a _maximum amplification_ due to either their demographic rates or the initial vector (because they never attenuate), some only have a _maximum attenuation_ (because they never amplify), and some have both.

[^2]: [Stott, I., Hodgson, D. J., & Townley, S. (2012) Functional Ecology, Methods in Ecology & Evolution, 3(5).](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/j.2041-210X.2012.00222.x)

The function `resilience` can compute each of the aforementioned metric, calculated for each matrix that was made at each time step of the period of the study.  

The parameters to specify are:

•	`listA`: List of square, primitive, irreducible, non-negative numeric matrices of any dimension.

•	`metrics`: Character string. The user can specify either one or several resilience metrics of their choice, The metrics have to be specified as either `"convt"` (for _convergence time_); `"dr"` (for _damping ratio_);`"maxamp"` (for _maximum amplification_); `"maxatt"` (for _maximum attenuation_); `"reac"` (for _reactivity_); or `"all"` to compute all of the aforementioned metrics. By default set to `"all"`.   

•	`vector`: Numeric. Can be a numeric vector or a list of numeric vectors describing the age/stage distribution ('demographic structure') used to calculate a 'case-specific' metric, i.e. based on the stage/age structure.            

•	`TDvector`: Boolean. Specifies whether or not the user wants to get a Time-Dependent list of initial vectors, corresponding to the population stage distribution projection for each year with each matrix of each year. The result is a list of X initial vectors with X the number of matrices that are in the list. By default set to `FALSE`.        

•	`popname`: Character string describing the name of the population.

•	`accuracy`: Numeric. This is an option taken from the function `convt` from __{popdemo}__ to use when calculating the _convergence time_. It is a numeric value between 0  and 1 representing the  accuracy with which to determine convergence to asymptotic growth, expressed as a proportion. By default set to 0.01.      

•	`iterations`: This is an option taken from the function `convt` from __{popdemo}__ to use when calculating the _convergence time_. It represents the maximum number of iterations of the model before the code breaks. By default set to 1e+05. 

•	`verbose`: Boolean. If set to `TRUE`, it will display the messages that are produced when assessing the demographic resilience metrics. By default set to `TRUE.`   

• `return.N`: Boolean. If set to `TRUE`, it will return the value of the population size at the point of the metric's value is reached. If set to `FALSE`, returns the standardized value f the metric. B default set to `TRUE`.

• `return.t`: Boolean. If set to `TRUE`, it will return the time at which the metric is reached. By default set to `FALSE`. 

## Calculating one resilience metric 
First, let us look at __{reactivity}__, by setting `metrics = "reac". We will first set return.N = FALSE to get the value of reactivity relative to the asymptotic growth rate and with a standardized vector. The function is taken from the __{popdemo}__ package and uses the same syntax. It automatically standardizes the Matrix (i.e. divides its entries by its asymptotic growth rate {\lambda}) and the initial vector (i.e. divides its entries by its sum to get proportions between 0 and 1).       

We will first create an initial vector representing the total number of inidividuals in each stage class. This is randomly generated for the sake of the example but the users can test different values of initial vectors (see section : "Initial vectors")

The initial vector should always be a list of vectors, with the length equal to the length of the list of matrices. They can be of different values, or of the same repeated value. 

Here we chose to repeat the same value of 21 juveniles and 39 adults. These numbers were chosen for the sake of illustration of the package and ond't have so far any biological implication or meaning.

```{r, reactivity standardized, echo = -1}
# simulate an initial vector
set.seed(125435)
vec <- round(runif(2) * 100 )
vector_list <- lapply(1:28, function(i) vec)

AP_demres_reac_stand <-
   resilience(
     listA = adeliepenguin,
     metrics = "reac",
     vector = vector_list,
     TDvector = FALSE,
     popname = "adelie penguin",
     verbose = TRUE,
     return.N = FALSE,
     return.t = TRUE
   )

AP_demres_reac_stand

```

We can see that __{reactivity}__ varies in time. The maximum values for the standardized calculation of __{reactivity}__, calculated with the provided vector was found during the 14th year of the study period. With a value of 1.0132911. 

A __{reactivity}__ of 1.01 would mean that in the first timestep, the population grows 1.01 as fast as a population that grows at its asymptotic growth rate. It is therefore slightly amplifying. However, we can also see that the year 14 is the only year for which the population amplifies. The rest of the years, the population attenuates (i.e. has a reactivity below 1) with the lowest value of reactivity reached in the 26th year, with a value of 0.91. 

## HERE plot of reactivity value for the 28 years

We can now look at which values take the corresponding population abundances after the first timestep. 

```{r, reactivity abundance, echo = -1}
# simulate an initial vector
set.seed(125435)
vec <- round(runif(2) * 100 )
vector_list <- lapply(1:28, function(i) vec)

AP_demres_reac <-
   resilience(
     listA = adeliepenguin,
     metrics = "reac",
     vector = vector_list,
     TDvector = FALSE,
     popname = "adelie penguin",
     verbose = TRUE,
     return.N = TRUE,
     return.t = TRUE
   )

AP_demres_reac
```

As for the standardized reactivity, the maximum population abundances for __{reactivity}__ calculated with the provided vector was found during the 14th year of the study period, with a value of 79.26 individuals. This is the projected number of indiviuals when the population reached its maximum reactivity. And the minimum value is also reached during year 26 with 52.14 individuals. 

## Calculating several metrics
Now let's look at all the metrics possible with the option `metrics = "all"`

```{r echo = -1}
AP_demres <-
   resilience(
     listA = adeliepenguin,
     metrics = "all",
     vector = vector_list,
     TDvector = FALSE,
     popname = "adelie penguin",
     verbose = TRUE,
     return.N = TRUE,
     return.t = TRUE
   )

AP_demres
```

This time we have an object of class __{resil}__ that is a dataframe with 11 columns. 

```{r echo = -1}
 colnames(AP_demres)
```

All 5 metrics are given and for all metrics but damping ratio we show the time when the metric is reached. Showing the time when reactivity is reached is somewhat redundant (as per definition it is the population size at time step 1) but we decided to display it nevertheless, for consistency. Messages were displayed due to the fact that for some combinations of specified matrices and vectors the population either does not amplify or does not attenuate. 

Given the population demographic structure specified in $\hat{\textbf{n}}_0$, it appears that the population amplifies in one of the study years only (year 14), in all other years it attenuates. Interestingly, for this year the reactivity value is exactly the same as that of maximum amplification (79.26) and maximum amplification is reached at time step 1. So, the population reachest its highest projected population size in the first time step and, in line with this, convergence time is 2 (with the same population size as for reactivity and maximum amplification). 

## Time-dependent initial vector
It is also possible to derive a list of initial vectors starting with one vector representing the demographic structure of the population at the first time step (option `TDvector = TRUE` for Time-Dependent vector). At the first time step, the population size is projected to the next time step using the specified MPM and the given initial vector (for demographic structure). This results in proportions of individuals projected to be in each stage/age, which is then used as the demographic structure for the next time step. This process continues for as many time steps as there are.      

From here on we will set `verbose = FALSE` to avoid the display of all the messages.

```{r echo = -1}
AP_demres_TD <-
   resilience(
     listA = adeliepenguin,
     metrics = "all",
     vector = vector_list,
     TDvector = TRUE,
     popname = "adelie penguin",
     verbose = FALSE,
     return.N = TRUE,
     return.t = TRUE
   )

AP_demres_TD
```

We can see that when using time-varying demographic vector while projecting the initial population vector forward, the population is more likely to amplify than it was when using the actually observed demographic structures for each year. We also see that the values of all resilience metrics are increasing rapidly over time and by the the year 28 we have population sizes in thousands as opposed to population sizes being below 100 when relying on the user-supplied population vectors. 

When using the option `TDvector = TRUE`, the function projects the population structure for each time step using the MPM of each time-step, which can be similar to projecting a population over time with one MPM except that here the function does it with a different matrix for each time step. It is therefore expected that the population will approach asymptotic growth over time as the function projects the population over each time step, which are translated with demographic resilience metrics getting closer to 1. **VIK: I am not sure this previous statement will hold now, as we change matrices and we do not standardise by anything. JUlie please check - if in doubt, better remove this statement** The variations that we can see therefore represent the variation in the demographic rates in the MPM in comparison with a projection over time with only one matrix. 


# Assess the variation in the time-varying resilience metrics 

The function `summary` calculates the distance between the resilience metrics calculated using the time-varying and the time-constant approach. The following commonly used distances can be measured: 
residual mean squared error: 
$$RMSE = \sqrt{mean((TV-TC)^2)}$$ 
relative residual mean squared error: 
$$rRMSE = \frac{\sqrt{mean((TV-TC)^2)}}{sd(TV)}$$ 
and the mean absolute proportional error:    
$$MAPE = \frac{mean(abs(TV-TC))}{TC}$$ 
with $TV$: the time-varying resilience metric and $TC$ the time constant one. 

## One variable
Let's look at how the metrics vary in time in comparison with the time-constant approach. Looking first at `reactivity`.
To be able to look at these distances, the metrics calculated with both time-varying and time-dependent approaches must be present in the table. 

```{r echo = -1}
# table.AP.reac <- summary(adeliepenguin.reac.tvtc)
# 
# table.AP.reac
```
The lower bound has the largest value of RMSE. This means that the upper bound is the one for which the time-varying values are the most spread around the time-constant value for _reactivity_. When dividing this value by the standard deviation of the time-varying metric, it seems that all three metrics have a rRMSE below 1, which means that the value of the RMSE is below the value of the SD of the time-varying metrics, which can be interpreted as a reasonable dispersion around the time-constant metric. Finally, when looking at the MAPE, if we take 5% as the maximum threshold above which the difference between the time-varying and the time-constant approaches are considered to be important, then using the time-varying approach seems to be sensible for computing the metric at the lower and upper bounds, whereas the time-constant approach is a good enough approximation for the value computed using the specified initial vector.  

## Several metrics
Then let's look at all the resilience metrics: 

```{r echo = -1}
# table.AP.all <- summary(adeliepenguin.all.tvtc)
# 
# table.AP.all
```
First, we can notice that the distance measures for _maximum attenuation_ cannot be computed because the population did not attenuate according to the time-constant approach. 

Looking at the RMSE, _damping ratio_ is the metric with the largest distance value. This is due to the fact that _damping ratio_ displays values varying between 3.21 and 8.61 while its value based on the average matrix was calculated to be 4.48. This show the high variability over time of he population's capacity to converge towards an equilibrium. The second metric with the largest value of RMSE is _convergence time_. This is due to the the fact that _convergence time_ is expressed in number of time steps, therefore the values can vary greatly, and are much larger than the values of the other metrics that are around 1. Secondly, when looking at the rRMSE, all other metrics' values seem to be below (but close to) 1, meaning that the dispersion of the metrics computed with the time-varying approach is within 1 SD of the time-constant approach. However, when looking at the MAPE, we see that MAPE has values above 5% for _convergence time_ calculated based on an initial vector and its bounds, _damping ratio_,  the lower bound of _inertia_, the lower bound of _maximum attenuation_, and the lower and upper bound of _reactivity_. For these resilience metrics, we can consider that the time-varying approach is more sensible than the time-constant approach.

It is now worth investigating the reasons behind the fact that the lower bound appears to be the metric that varies the most in time. 

# Check the stage-biased vectors 
Let's first remember the initial vector we used 

```{r echo = -1}
# penguinvec1
```
For this initial vector, adults were largely represented, with juveniles being in minority. 
Now we will create the 2 stage-biased vectors of our population and identify the cases in which the upper and lower bounds are found:

```{r echo = -1}
## creating the stage-biased vectors
# stage1 <- c(1,0)
# stage2 <- c(0,1)
# 
# list.initvec <- list(stage1,
#                      stage2)
# 
# list.demres <- lapply(X = list.initvec, 
#                       FUN = resilience, 
#                       listA = adeliepenguin,
#                       popname = "adeliepenguin",
#                       bounds = TRUE,
#                       metrics = "all",
#                       verbose = FALSE)
# 
# 
# list.demres
```

If there is an over-representation of individuals in stages with low survival and zero/low fertility, then the population will grow slower (or decline faster) than the stable rate (i.e. attenuates). Such a case is represented with the initial vector `stage 1` (that can be seen by computing `list.demres[[1]]`) above, which seems to be the one that leads to the lowest values of _inertia_, _maximum attenuation_, _reactivity_ and the upper bounds of _convergence time_. This means that if the population consists of juveniles only: the population will attenuate to its lowest possible value and will take the longest time to converge. 

If there is an over-representation of individuals in stages with high survival and/or fertility, then the population will grow faster (or decline slower) than the stable rate (i.e. amplifies). Such a case is represented with the initial vector `stage 2` (that can be seen by computing `list.demres[[2]]`)above, which seems to be the one that leads to the highest values of _inertia_, _maximum attenuation_, _reactivity_ and the lower bounds of _convergence time_. This means that if the population consists of reproducing adults only: the population will amplify to its highest possible value and will take the shortest time to converge. 
    
As we previously saw, the lower bound is the one with the largest values of distance between the time-varying and the time-constant approach. This means that the lower bound seems to vary the most in time in comparison to the other demographic resilience metrics. We saw that the lower bound of the demographic resilience metrics (except _convergence time_) can be obtained when the initial vector is biased towards juveniles only. The high variation in the lower bounds are therefore a representation of the variation in time of the demographic rates of the juveniles. 

# comparisn when the vector is time-dependent
```{r echo = -1}
# adeliepenguin.all.tvtc.TD <- resilience(
#      listA = adeliepenguin,
#      metrics = "all",
#      bounds = TRUE,
#      vector = penguinvec1,
#      TDvector = TRUE,
#      popname = "adelie penguin",
#      time = "both",
#      verbose = FALSE)
# 
# table.AP.all.TD <- summary(adeliepenguin.all.tvtc.TD)
# 
# table.AP.all.TD

```

In the case of a time-dependent initial vector, the distance measures are different. _daping ratio_ is still the metric that shows the highest values of RMSE, however the rRMSE values this time indicate that the _convergence time_ and _maximum amplification_ both have a RMSE larger than a single standard deviation calculated across the values obtained with time-varying approach. This indicates a broad dispersion of the metrics obtained with the time-varying approach around the value obtained using the time-constant approach. 
Finally, when looking at the MAPE, almost all metrics this time show a MAPE above 5% except for _inertia_ and its upper bound.  
Values of _damping ratio_ are not changing in that case because it is calculated based on the properties of the matrix and does not depend on the initial vector. 

When initial vectors depend on time, we can conclude that the time-varying approach is even more sensible than when using only one initial vector across the time-steps. 

# Check extreme years 
We have seen already that the years 4,9,12 and 15 were the best in terms of adult survival probability, while he year 21 was the best for transition probability and the year 14 was the best in terms of fecundity. 
On the other hand, the year 22 was the worst in terms of adult survival, the year 10 was he worst in terms of transition probability and the ear 7 was the worst for the fecundity.

Let's see how this translates in terms of demographic resilience metrics. 
## HERE 
### Maximum values 
```{r echo = -1}
#maximum values 
# apply(adeliepenguin.all.tv[-c(1:2)], 2, which.max)

```

The second year is the year during which most of the demographic resilience metrics are at their maximum value. The population amplifies, given the initial vector that was provided, however the extent of the amplification remains close to 1, with _inertia_ of 1.6, a _maximum amplification_ of 1.9 and a _reactivity_ close to 1.4. The maximum the population amplifies at is around 3.7. However, the second year also corresponds to the highest value of _convergence time_ and the lowest value of _damping ratio_. Given that the closer to 1 the _damping ratio_ is, the slower the population converges, we can conclude that during the second year, the population reaches maximum values of amplification but converges the slowest. 

### Minimum values 
```{r echo = -1}
#minimum values 
# apply(adeliepenguin.all.tv[-c(1:2)], 2, which.min)
```

Now looking at the minimum values, we can observe that the 5th year was the year during which the demographic resilience metrics were at their lowest, however with _convergence time_ at its lowest value and _damping ratio_ at its highest value, displaying a trade-off between amplification and convergence (resistance and recovery in other words). 

# Plot 
The function `plot` provides a visual representation of the time-varying and time-constant metrics that are present in the data frame returned by the function `resilience`. Depending on what the user specified as demographic distribution, the plot shows the resilience metrics computed by using  the lower bound, the upper bound, or the initial distribution vector. 

```{r, echo = -1, fig.width = 6, fig.height = 6}
# plot(adeliepenguin.all.tvtc)
```

R opened as many windows as there are metrics in the table provided. In each window is a visual representation of the demographic resilience metrics over time. 

