---
title: "demres: quantification of demographic resilience over time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{demres: quantification of demographic resilience over time}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "##  ", collapse = TRUE, fig.align = "center")
library(demres)
options(digits = 4)
```

The goal of __{demres}__ is to provide easy functions to calculate different time-varying and time-constant demographic resilience metrics. It also allows plotting the resulting metrics and measuring the discrepancy between the time-varying and the time-constant approach. 

The different metrics provided are:   
- _Convergence time_    
- _Damping ratio_    
- _Inertia_  
- _Maximum amplification_   
- _Maximum attenuation_   
- _Reactivity_   

It is build around one dependency:

- [__{popdemo}__](https://github.com/iainmstott/popdemo)

__{demres}__ aims at being compatible with both _tidyverse_ and _base_ R dialects.

***
# Installing `demres`
__{demres}__ is available on [GitHub](https://github.com/JulieLouvrier/demres). 
Vignette description usually requires the latest version of the package. 

## Package installation from Github

You can install this package using __{remotes}__ (or __{devtools}__):

```{r, installation, eval = FALSE}
# Install dependencies from CRAN:
install.packages(c("devtools", "remotes"))

#Installing from GitHub
remotes::install_github("JulieLouvrier/demres")
```
***  
# Matrix Population Models
## adelie penguin models

We will use a list of matrix projection models (MPMs) of an Adelie penguin population (*Pygoscelis adeliae*) from the Copacabana colony on King George Island, Antarctica, from the study by Hinke et al. (2017): "Variable vital rates and the risk of population declines in Adelie
penguins from the Antarctic Peninsula region" [^1].

The MPMs were extracted from the COMADRE database (https://compadre-db.org/). The data is a list of 28 matrices of dimension 2 x 2. The 2 stage classes have been defined based on age and are the following:     
- "Juveniles" - ages 0–2      
- "Adults" - ages 3+         
   

[^1]: [Hinke et al. 2017 Ecosphere, 8(1), e01666.](https://esajournals.onlinelibrary.wiley.com/doi/epdf/10.1002/ecs2.1666)

```{r echo = FALSE}
penguinpic <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/1/12/Pygoscelis_adeliae.png")
par(mar = c(0, 0, 0, 0))
plot(as.raster(penguinpic))
text(10, 10, adj = c (0, 0), col = "white",
     "Adelie penguin")
```

Load in the data:  
```{r}
data(adeliepenguin); adeliepenguin

```

#####REPRENDRE ICI AUSSI
The numbers in the matrices describe the elements known as vital rates. These represent the survival rates in a given (st)age class, transition probabilities from one (st)age class to another, and fecundity (i.e. per-capita number of offspring that is contributed to the population by each (st)age class), per unit of time. For Adelie penguins, the highest value of fecundity (0.75) was found in year 14 of the study (see `adeliepenguin[[14]]`), while this parameter reached its lowest value of 0.16 in year 7 of the study period (see `adeliepenguin[[7]]`). The adult survival rate varied across years with its highest value (0.98) on years 4, 9, 12 and 15 and lowest values (0.72) on the year 22. The probability to transition from a juvenile to an adult reached its highest value (0.89) on year 21 and its lowest value (0.53) on years 10 and 28. 


# Assess time-varying demographic resilience 
The function `resilience` is made to calculate demographic resilience metrics based on a list of MPMs. It returns a `dataframe` with time-varying demographic resilience metrics with the option `time = "varying"`.

The different resilience metrics that are calculated[^2] can be :         

- _**Convergence time**_ : Time to convergence of a population from the model projection.         

- _**Damping ratio**_ : Dimensionless measure of convergence to stable growth. Smaller numbers represent slower convergence. _Damping ratio_ is calculated using 

$$\textit{damping ratio} = {\rho} = \frac{\lambda_1}{||\lambda_2||}$$
With $\lambda_1$ the dominant eigenvalue and $\lambda_2$ the largest subdominant eigenvalue.

- _**Inertia**_ : The long-term population density in comparison with a population that reached a stable growth. _Inertia_ is calculated using  

$$\textit{inertia} = {P}_{\infty} = \frac{\textbf{v}^\text{T}\hat{\textbf{n}}_0||\textbf{w}||_1}{\textbf{v}^\text{T}\textbf{w}}$$ 

in case of an amplifying population and 

$$\textit{inertia} = {\rho}_{\infty} = \frac{\textbf{v}_{min}||\textbf{w}||_1}{\textbf{v}^\text{T}\textbf{w}}$$ 
in case of an attenuating population. 

With **w** and **v** being right and left eigenvectors respectively. $\textbf{v}_{min}$ and $\textbf{v}_{max}$ represent the smallest and largest entries of **v**. $||\textbf{w}||_1$ represents the one-norm of **w** (equals to the sum of its entries). 

- _**Reactivity**_ : The population density that can be reached in the first time step after disturbance. _Reactivity_ is calculated using  

$$\textit{reactivity} = {P_1} = ||\hat{\textbf{A}} \hat{\textbf{n}}_0||_1$$
With **A** the matrix population model and $\hat{\textbf{A}}$ the standardized population matrix which is calculated as $\hat{\textbf{A}} = \frac{\textbf{A}}{\lambda_{1}}$. The vector $\hat{\textbf{n}}_0$ represents the initial demographic distribution, standardized to sum to 1 and is calculated as $\hat{\textbf{n}}_0 = \frac{{\textbf{n}}_0}{||\textbf{n}_0||_1}$. 

- _**Maximum amplification**_ : The largest population density that can be reached at any time after disturbance. This metric is calculated using:
$$maximum\ amplification = \bar{P}_{max} = \max_{t > 0}||\hat{\textbf{A}}^t\hat{\textbf{n}}_0||_1$$

- _**Maximum attenuation**_ : The lowest population density that can be reached at any time after disturbance. This metric is calculated using: 

$$maximum\ attenuation = \underline{P}_{min} = \min_{t > 0}||\hat{\textbf{A}}^t\hat{\textbf{n}}_0||_1$$
_Maximum amplification_ and _maximum attenuation_ can occur at any point along the projection. Some populations only have a _maximum amplification_ due to either their demographic rates or the initial vector (because they never attenuate), some only have a _maximum attenuation_ (because they never amplify), and some have both.

[^2]: [Stott, I., Hodgson, D. J., & Townley, S. (2012) Functional Ecology, Methods in Ecology & Evolution, 3(5).](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/j.2041-210X.2012.00222.x)

The function `resilience` can compute each of the aforementioned metric, calculated for each matrix that was made at each time step of the period of the study.  

The parameters to specify are:

•	`listA`: List of square, primitive, irreducible, non-negative numeric matrices of any dimension.

•	`metrics`: Character string. The user can specify either one or several resilience metrics of their choice, The metrics have to be specified as either `"convt"` (for _convergence time_), `"dr"` (for _damping ratio_); `"inertia"`, `"maxamp"` (for _maximum amplification_); `"maxatt"` (for _maximum attenuation_); `"reac"` (for _reactivity_); or `"all"` to compute all of the aforementioned metrics. By default set to `"all"`.             

•	`bounds`: Boolean. Optional argument if the user wants to calculate the upper and lower bounds calculated from the stage-biased vectors. By default set to `TRUE`.         

If `bounds = TRUE` but no `vector` is specified, the function provides metrics at their upper and lower bound, calculated based on the stage-biased vector. If `vector` is specified, the function additionally provides the metrics calculated based on the provided initial vector.

•	`vector`: Numeric. Can be a numeric vector or a list of numeric vectors describing the age/stage distribution ('demographic structure') used to calculate a 'case-specific' metric, i.e. based on the stage/age structure.            

•	`TDvector`: Boolean. Specifies whether or not the user wants to get a Time-Dependent list of initial vectors, corresponding to the population stage distribution projection for each year with each matrix of each year. The result is a list of X initial vectors with X the number of matrices that are in the list. By default set to `FALSE`.        

•	`popname`: Character string describing the name of the population.      

•	`time`: Character string: `"constant"` (if the metrics are to be calculated over the whole study period); `"varying"` (if the metrics are to be calculated for each time step) or `"both"`.          
•	`accuracy`: Numeric. This is an option taken from the function `convt` from __{popdemo}__ to use when calculating the _convergence time_. It is a numeric value between 0  and 1 representing the  accuracy with which to determine convergence to asymptotic growth, expressed as a proportion. By default set to 0.01.      
•	`iterations`: This is an option taken from the function `convt` from __{popdemo}__ to use when calculating the _convergence time_. It represents the maximum number of iterations of the model before the code breaks. By default set to 1e+05. 

•	`verbose`: Boolean. If set to `TRUE`, it will display the messages that are produced when assessing the demographic resilience metrics. By default set to `TRUE.`        

## Calculating one resilience metric 
First, let us look at _inertia_, by setting `metrics = "inertia"` and `time = "varying"`.

```{r echo = -1}
set.seed(1234)
penguinvec1 <- runif(2)
penguinvec1 <- penguinvec1 / sum(penguinvec1)

adeliepenguin.inertia.tv <- resilience(
  listA = adeliepenguin,
  metrics = "inertia",
  bounds = TRUE,
  vector = penguinvec1,
  popname = "adelie penguin",
  time = "varying"
)

adeliepenguin.inertia.tv
```

Here we can see that _inertia_ varies in time. Since `bounds` were set to `TRUE`, the boundary values of _inertia_ were calculated also based on the stage-biased vector. Stage-biased vectors are biased towards one stage class, with individuals from this stage class representing 100% of the whole population. Consequently there are as many stage-biased vectors as there are stage classes in the population. For example, a population with 4 stage classes will have the 4 following stage-biased vectors: [1 0 0 0]; [0 1 0 0]; [0 0 1 0]; [0 0 0 1 ]. These upper and lower bounds therefore provide the upper and lower extremes that the demographic resilience metrics can take. 


The minimum values for _inertia_ for its upper bound and calculated with the vector were both found during the 7th year of the study period (1 and 1.05 respectively), while it was found during the 10th year of the study for the lower bound of inertia (0.54). While its upper bound's maximum (1.25) and maximum value based on the initial vector we provided (1.14) were both reached on the 14th year. While the maximum value of the lower bound of inertia was found on the 21st year of the study (0.80). An _inertia_ of 1.25 means that after the transient period, the population would settle to a size 1.25 times as large as a population that grows with stable rate.

Second, let us look at _reactivity_ by setting `metrics = "reac"`:
```{r echo = -1}
adeliepenguin.reac.tv <- resilience(
     listA = adeliepenguin,
     metrics = "reac",
     bounds = TRUE,
     vector = penguinvec1,
     popname = "adelie penguin",
     time = "varying")

adeliepenguin.reac.tv
```

A _reactivity_ of 1.32 would mean that in the first timestep, the population grows 1.32 as fast as a population that grows at its stable growth rate. It is therefore amplifying. 

## Calculating several metrics
Now let's look at all the metrics possible with the option `metrics = "all"`

```{r echo = -1}
adeliepenguin.all.tv <- resilience(
     listA = adeliepenguin,
     metrics = "all",
     bounds = TRUE,
     vector = penguinvec1,
     popname = "adelie penguin",
     time = "varying")

adeliepenguin.all.tv
```

This time we have an object of class __resil__ that is a dataframe with 16 columns. 

```{r echo = -1}
colnames(adeliepenguin.all.tv)
```

All 6 metrics are given with their upper and lower bounds when possible. Messages were displayed due to the fact that the bounds were requested for _maximum amplification_ and _maximum attenuation_, but the lower bound cannot be calculated for _maximum amplification_ and the upper bound cannot be calculated for _maximum attenuation_. 

The lower and upper bounds are calculated based on the stage-biased vector, therefore the function warn the users to interpret the results with caution when looking at the lower and upper bounds of _convergence time_. _Convergence time_ represents the number of time-steps it will take for the population to reach asymptotic growth rate. Therefore, the stage-biased vectors might lead to longer _convergence time_ than when calculated with an initial vector representing more realistic demographic structure. 

Given the population demographic structure specified in $\hat{\textbf{n}}_0$, it appears that for none of the years the population would attenuate, and rather amplifies.

## Time-dependent initial vector
It is also possible to derive a list of initial vectors starting with one vector representing the demographic structure of the population at the first time step (option `TDvector = TRUE` for Time-Dependent vector). At the first time step, the population size is projected to the next time step using the specified MPM and the given initial vector (for demographic structure). This results in proportions of individuals projected to be in each stage/age, which is then used as the demographic structure for the next time step. This process continues for as many time steps as there are.      

If `time = "both"` the time-constant demographic resilience metrics are calculated based on the averaged vector over the years.   

From here on we will set `verbose = FALSE` to avoid the display of all the messages.

```{r echo = -1}
adeliepenguin.all.tv.TD <- resilience(
     listA = adeliepenguin,
     metrics = "all",
     bounds = TRUE,
     vector = penguinvec1,
     TDvector = TRUE,
     popname = "adelie penguin",
     time = "varying",
     verbose = FALSE)

adeliepenguin.all.tv.TD
```
Here we can see that values differ from the previous `adeliepenguin.all.tv`. Indeed, _inertia_ and _reactivity_ often have values that are below 1 when the vector provided is time-dependent, meaning that the population is attenuating rather than amplifying during these years.  

In comparison with the previous table, one can see that the values of _inertia_ and _reactivity_, calculated with the time-dependent initial vector converge towards 1 over time. 

We can see this by comparing _inertia_ when the initial vector is static
```{r}
adeliepenguin.all.tv$inertia_TV
```

with values of _inertia_ when the initial vector is time-dependent
```{r}
adeliepenguin.all.tv.TD$inertia_TV
```
Let's take a look at what _inertia_ looks like when using a time-dependent initial vector in comparison with one non-changing value of initial vector. 

```{r echo = -1}
adeliepenguin.all.tv$timestep <- factor(adeliepenguin.all.tv$timestep) 
                                    
plot(adeliepenguin.all.tv$inertia_TV, ylim = c(0.6, 1.3), type = "l", 
     xlab = "Time-step",
     ylab = "Inertia of adelie penguins")
lines(adeliepenguin.all.tv.TD$inertia_TV, col = "red")
legend(1.5, 0.9, legend = c("Time dependent", "No time dependent"),
       col = c("red", "black"), cex = 0.6, lty = c(1, 1), box.lty = 0)
abline(h = 1, col = "blue", lty = 2)
```

Let's look now at _reactivity_:

```{r echo = -1}
adeliepenguin.all.tv$timestep <- factor(adeliepenguin.all.tv$timestep) 
                                    
plot(adeliepenguin.all.tv$reac_TV, ylim = c(0.6, 1.3), type = "l", 
     xlab = "Time-step",
     ylab = "Reactivity of adelie penguins")
lines(adeliepenguin.all.tv.TD$reac_TV, col = "red")
legend(1.5, 0.9, legend = c("Time dependent", "No time dependent"),
       col = c("red", "black"), cex = 0.6, lty = c(1, 1), box.lty = 0)
abline(h = 1, col = "blue", lty = 2)
```

When using the option `TDvector = TRUE`, the function projects the population structure for each time step using the MPM of each time-step, which can be similar to projecting a population over time with one MPM except that here the function does it with a different matrix for each time step. It is therefore expected that the population will approach asymptotic growth over time as the function projects the population over each time step, which are translated with demographic resilience metrics getting closer to 1. The variations that we can see therefore represent the variation in the demographic rates in the MPM in comparison with a projection over time with only one matrix. 

Note that it is also possible to provide a list of initial vectors, of a length that is equal to the number of MPMs present in `listA`. This can be used when users have information about the actual demographic structure of their study population over time. 

# Assess time-constant demographic resilience
## Using one resilience metric
First we can look at _inertia_ again:
```{r echo = -1}
adeliepenguin.inertia.tc <- resilience(
     listA = adeliepenguin,
     metrics = "inertia",
     bounds = TRUE,
     vector = penguinvec1,
     popname = "adelie penguin",
     time = "constant",
     verbose = FALSE)

adeliepenguin.inertia.tc
```
Here we can see that _inertia_ is calculated based on the mean MPM and is around 1.07 with a lower bound of 0.63 and an upper bound of 1.15 

## Using several resilience metrics
We can look now at all metrics with the option `metrics = "all"`:
```{r echo = -1}
adeliepenguin.all.tc <- resilience(
     listA = adeliepenguin,
     metrics = "all",
     bounds = TRUE,
     vector = penguinvec1,
     popname = "adelie penguin",
     time = "constant",
     verbose = FALSE)

adeliepenguin.all.tc
```

The population does not attenuate based on the average matrix and the initial vector that was provided.

# Compare time-varying with time-constant approaches 
## One resilience metric
```{r echo = -1}
adeliepenguin.reac.tvtc <- resilience(
     listA = adeliepenguin,
     metrics = "reac",
     bounds = TRUE,
     vector = penguinvec1,
     popname = "adelie penguin",
     time = "both", 
     verbose = FALSE)

adeliepenguin.reac.tvtc
```
Both time-varying and time-constant approaches are displayed in the table for _reactivity_: time-constant ones have `_TC` appended in the end of the name, whereas time-varying ones have `_TV`.

## Several metrics
```{r echo = -1}
adeliepenguin.all.tvtc <- resilience(
     listA = adeliepenguin,
     metrics = "all",
     bounds = TRUE,
     vector = penguinvec1,
     popname = "adelie penguin",
     time = "both", 
     verbose = TRUE)

adeliepenguin.all.tvtc

```
Because `verbose = TRUE` we can see the messages displayed for each time step of the time varying approach, as well as the messages for the time-constant approach. 

# Assess the distance between the time-varying and the time-constant approaches 

The function `summary` calculates the distance between the resilience metrics calculated using the time-varying and the time-constant approach. The following commonly used distances can be measured: 
residual mean squared error: 
$$RMSE = \sqrt{mean((TV-TC)^2)}$$ 
relative residual mean squared error: 
$$rRMSE = \frac{\sqrt{mean((TV-TC)^2)}}{sd(TV)}$$ 
and the mean absolute proportional error:    
$$MAPE = \frac{mean(abs(TV-TC))}{TC}$$ 
with $TV$: the time-varying resilience metric and $TC$ the time constant one. 

## One variable
Let's look at how the metrics vary in time in comparison with the time-constant approach. Looking first at `reactivity`.
To be able to look at these distances, the metrics calculated with both time-varying and time-dependent approaches must be present in the table. 

```{r echo = -1}
table.AP.reac <- summary(adeliepenguin.reac.tvtc)

table.AP.reac
```
The lower bound has the largest value of RMSE. This means that the upper bound is the one for which the time-varying values are the most spread around the time-constant value for _reactivity_. When dividing this value by the standard deviation of the time-varying metric, it seems that all three metrics have a rRMSE below 1, which means that the value of the RMSE is below the value of the SD of the time-varying metrics, which can be interpreted as a reasonable dispersion around the time-constant metric. Finally, when looking at the MAPE, if we take 5% as the maximum threshold above which the difference between the time-varying and the time-constant approaches are considered to be important, then using the time-varying approach seems to be sensible for computing the metric at the lower and upper bounds, whereas the time-constant approach is a good enough approximation for the value computed using the specified initial vector.  

## Several metrics
Then let's look at all the resilience metrics: 

```{r echo = -1}
table.AP.all <- summary(adeliepenguin.all.tvtc)

table.AP.all
```
First, we can notice that the distance measures for _maximum attenuation_ cannot be computed because the population did not attenuate according to the time-constant approach. 

Looking at the RMSE, _damping ratio_ is the metric with the largest distance value. This is due to the fact that _damping ratio_ displays values varying between 3.21 and 8.61 while its value based on the average matrix was calculated to be 4.48. This show the high variability over time of he population's capacity to converge towards an equilibrium. The second metric with the largest value of RMSE is _convergence time_. This is due to the the fact that _convergence time_ is expressed in number of time steps, therefore the values can vary greatly, and are much larger than the values of the other metrics that are around 1. Secondly, when looking at the rRMSE, all other metrics' values seem to be below (but close to) 1, meaning that the dispersion of the metrics computed with the time-varying approach is within 1 SD of the time-constant approach. However, when looking at the MAPE, we see that MAPE has values above 5% for _convergence time_ calculated based on an initial vector and its bounds, _damping ratio_,  the lower bound of _inertia_, the lower bound of _maximum attenuation_, and the lower and upper bound of _reactivity_. For these resilience metrics, we can consider that the time-varying approach is more sensible than the time-constant approach.

It is now worth investigating the reasons behind the fact that the lower bound appears to be the metric that varies the most in time. 

# Check the stage-biased vectors 
Let's first remember the initial vector we used 

```{r echo = -1}
penguinvec1
```
For this initial vector, adults were largely represented, with juveniles being in minority. 
Now we will create the 2 stage-biased vectors of our population and identify the cases in which the upper and lower bounds are found:

```{r echo = -1}
## creating the stage-biased vectors
stage1 <- c(1,0)
stage2 <- c(0,1)

list.initvec <- list(stage1,
                     stage2)

list.demres <- lapply(X = list.initvec, 
                      FUN = resilience, 
                      listA = adeliepenguin,
                      popname = "adeliepenguin",
                      bounds = TRUE,
                      metrics = "all",
                      verbose = FALSE)


list.demres
```

If there is an over-representation of individuals in stages with low survival and zero/low fertility, then the population will grow slower (or decline faster) than the stable rate (i.e. attenuates). Such a case is represented with the initial vector `stage 1` (that can be seen by computing `list.demres[[1]]`) above, which seems to be the one that leads to the lowest values of _inertia_, _maximum attenuation_, _reactivity_ and the upper bounds of _convergence time_. This means that if the population consists of juveniles only: the population will attenuate to its lowest possible value and will take the longest time to converge. 

If there is an over-representation of individuals in stages with high survival and/or fertility, then the population will grow faster (or decline slower) than the stable rate (i.e. amplifies). Such a case is represented with the initial vector `stage 2` (that can be seen by computing `list.demres[[2]]`)above, which seems to be the one that leads to the highest values of _inertia_, _maximum attenuation_, _reactivity_ and the lower bounds of _convergence time_. This means that if the population consists of reproducing adults only: the population will amplify to its highest possible value and will take the shortest time to converge. 
    
As we previously saw, the lower bound is the one with the largest values of distance between the time-varying and the time-constant approach. This means that the lower bound seems to vary the most in time in comparison to the other demographic resilience metrics. We saw that the lower bound of the demographic resilience metrics (except _convergence time_) can be obtained when the initial vector is biased towards juveniles only. The high variation in the lower bounds are therefore a representation of the variation in time of the demographic rates of the juveniles. 

# comparisn when the vector is time-dependent
```{r echo = -1}
adeliepenguin.all.tvtc.TD <- resilience(
     listA = adeliepenguin,
     metrics = "all",
     bounds = TRUE,
     vector = penguinvec1,
     TDvector = TRUE,
     popname = "adelie penguin",
     time = "both",
     verbose = FALSE)

table.AP.all.TD <- summary(adeliepenguin.all.tvtc.TD)

table.AP.all.TD

```

In the case of a time-dependent initial vector, the distance measures are different. _daping ratio_ is still the metric that shows the highest values of RMSE, however the rRMSE values this time indicate that the _convergence time_ and _maximum amplification_ both have a RMSE larger than a single standard deviation calculated across the values obtained with time-varying approach. This indicates a broad dispersion of the metrics obtained with the time-varying approach around the value obtained using the time-constant approach. 
Finally, when looking at the MAPE, almost all metrics this time show a MAPE above 5% except for _inertia_ and its upper bound.  
Values of _damping ratio_ are not changing in that case because it is calculated based on the properties of the matrix and does not depend on the initial vector. 

When initial vectors depend on time, we can conclude that the time-varying approach is even more sensible than when using only one initial vector across the time-steps. 

# Check extreme years 
We have seen already that the years 4,9,12 and 15 were the best in terms of adult survival probability, while he year 21 was the best for transition probability and the year 14 was the best in terms of fecundity. 
On the other hand, the year 22 was the worst in terms of adult survival, the year 10 was he worst in terms of transition probability and the ear 7 was the worst for the fecundity.

Let's see how this translates in terms of demographic resilience metrics. 
## HERE 
### Maximum values 
```{r echo = -1}
#maximum values 
apply(adeliepenguin.all.tv[-c(1:2)], 2, which.max)

```

The second year is the year during which most of the demographic resilience metrics are at their maximum value. The population amplifies, given the initial vector that was provided, however the extent of the amplification remains close to 1, with _inertia_ of 1.6, a _maximum amplification_ of 1.9 and a _reactivity_ close to 1.4. The maximum the population amplifies at is around 3.7. However, the second year also corresponds to the highest value of _convergence time_ and the lowest value of _damping ratio_. Given that the closer to 1 the _damping ratio_ is, the slower the population converges, we can conclude that during the second year, the population reaches maximum values of amplification but converges the slowest. 

### Minimum values 
```{r echo = -1}
#minimum values 
apply(adeliepenguin.all.tv[-c(1:2)], 2, which.min)
```

Now looking at the minimum values, we can observe that the 5th year was the year during which the demographic resilience metrics were at their lowest, however with _convergence time_ at its lowest value and _damping ratio_ at its highest value, displaying a trade-off between amplification and convergence (resistance and recovery in other words). 

# Plot 
The function `plot` provides a visual representation of the time-varying and time-constant metrics that are present in the data frame returned by the function `resilience`. Depending on what the user specified as demographic distribution, the plot shows the resilience metrics computed by using  the lower bound, the upper bound, or the initial distribution vector. 

```{r, echo = -1, fig.width = 6, fig.height = 6}
plot(adeliepenguin.all.tvtc)
```

R opened as many windows as there are metrics in the table provided. In each window is a visual representation of the demographic resilience metrics over time. 

